---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# eulerr

[![Travis-CI Build Status](https://travis-ci.org/jolars/eulerr.svg?branch=master)](https://travis-ci.org/jolars/eulerr)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/jolars/eulerr?branch=master&svg=true)](https://ci.appveyor.com/project/jolars/eulerr)
[![codecov](https://codecov.io/gh/jolars/eulerr/branch/master/graph/badge.svg)](https://codecov.io/gh/jolars/eulerr)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/eulerr)](https://cran.r-project.org/package=eulerr)

**eulerr** generates area-proportional euler diagrams that display set
relationships (intersections, unions, and disjoints) with circles.
[Euler diagrams](https://en.wikipedia.org/wiki/Euler_diagram) are Venn
diagrams without the requirement that all set interactions be present (whether
they are empty or not). That is, depending on input, eulerr will sometimes
produce Venn diagrams but sometimes not.

With three or more sets intersecting, exact euler diagrams are often
impossible. For such cases eulerr attempts to provide the best
approximation possible by numerically tuning the circles' positions and radiuses
so that the sum of squared errors is minimized.

When solutions are approximate, residuals and error statistics are provided
to assess whether the resulting diagram can be trusted.

## Installation

The CRAN version can be installed by running

```{r install_cran, eval = FALSE}
install.packages("eulerr")
```

The development version can be installed by running

```{r install_github, eval = FALSE}
devtools::install_github("jolars/eulerr")
```

## Usage

```{r sets}
library(eulerr)
sets <- c("A" = 10, "B" = 9, "C" = 4,
          "A&B" = 2, "A&C" = 3, "B&C" = 3,
          "A&B&C" = 2)
set.seed(1)
fit <- eulerr(sets)
```

We can inspect the solution

```{r}
fit
```

and plot it using `plot`.

```{r plot_methods, eval = FALSE}
plot(fit)
```

Please see the introductory [vignette](https://CRAN.R-project.org/package=eulerr/vignettes/introduction.html)
for usage details.

## License

eulerr is open source software, licensed under [GPL-3](LICENSE).

## Code of conduct

Please note that this project is released with a
[Contributor Code of Conduct](CONDUCT.md). By participating in this project you
agree to abide by its terms.

## Thanks

eulerr would not be possible without Ben Fredrickson's work on
[venn.js](http://www.benfrederickson.com) or Leland Wilkinson's
[venneuler](https://cran.r-project.org/package=venneuler).
