---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# eulerr

[![Travis-CI Build Status](https://travis-ci.org/jolars/eulerr.svg?branch=master)](https://travis-ci.org/jolars/eulerr)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/jolars/eulerr?branch=master&svg=true)](https://ci.appveyor.com/project/jolars/eulerr)
[![codecov](https://codecov.io/gh/jolars/eulerr/branch/master/graph/badge.svg)](https://codecov.io/gh/jolars/eulerr)

eulerr generates area-proportional euler diagrams that display set
relationships (intersections, unions, and disjoints) using circles.
[Euler diagrams](https://en.wikipedia.org/wiki/Euler_diagram) are Venn
diagrams without the requirement that all set interactions be present (whether
they are empty or not). Thus, depending on input, eulerr will sometimes
produce Venn diagrams but other times not.

With three or more sets interacting, exact euler diagrams are frequently
impossible. In these cases, eulerr will provide the best
approximation possible by numerically tuning the circles positions and radiuses
so that the sum of absolute errors is minimized.

When solutions are approximate, eulerr helpfully provides statistics
that allow the user decide if the approximation can be trusted.

## Installation

The development version can be installed by running

```{r install_github, eval = FALSE}
devtools::install_github("jolars/eulerr")
```

## Usage

```{r sets}
library(eulerr)
sets <- c("A" = 10, "B" = 9, "C" = 4,
          "A&B" = 2, "A&C" = 3, "B&C" = 3,
          "A&B&C" = 2)
fit <- eulerr(sets)
```

We look at the solution,

```{r}
fit
```

and plot it using `plot`.

```{r plot_methods, eval = FALSE}
plot(fit)
```

## License

eulerr is open source software, licensed under GPL-3.

## Thanks

eulerr would not be possible without Ben Fredrickson's work on
[venn.js](http://www.benfrederickson.com) or Leland
Wilkinson's
[venneuler](https://cran.r-project.org/web/packages/venneuler/index.html).
