---
title: "A Gallery of Euler and Venn Diagrams"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Gallery of Euler and Venn Diagrams}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.height = 4,
  fig.width = 5
)
```

```{r setup}
library(eulerr)
```

This vignette serves as a gallery for Euler diagrams and
as a showcase of the various options for customization that are available.

## Euler Diagrams

### Uniform intersections

```{r}
uniform_intersections <- euler(c("A" = 10, "B" = 10, "C" = 10,
                                 "A&B" = 4, "A&C" = 4, "B&C" = 4,
                                 "A&B&C" = 2))
plot(uniform_intersections)
```

### Disjoint sets

```{r}
disjoint_sets <- euler(c(A = 1, B = 1, C = 1))
plot(disjoint_sets, 
     labels = c("Tom", "Greg", "Alberta"),
     edges = list(lty = 1:3))
```

### A set contained in the intersection of two sets

```{r}
completely_contained <- euler(c("A" = 15, "B" = 15, "C" = 0,
                                "A&B" = 3, "A&C" = 0, "B&C" = 0,
                                "A&B&C" = 3))
plot(completely_contained,
     labels = list(col = c("white", "black", "black")),
     edges = list(col = "white", lex = 2),
     fills = c("black", "cyan", "orange"))
```

### Two sets intersecting inside a third

```{r}
intersecting_inside <- euler(c("A" = 15, "B" = 0, "C" = 0,
                               "A&B" = 3, "A&C" = 3, "B&C" = 0,
                                "A&B&C" = 2))
plot(intersecting_inside,
     fills = list(fill = c("lavenderblush2", 
                           "lightblue2", 
                           "lightsalmon",
                           "", 
                           "", 
                           "",
                           "plum2")),
     legend = list(side = "right"))
```

### Difficult set (for circles!)

```{r}
one_contained <- euler(c("A" = 7, B = 6, C = 0, 
                         "A&B" = 0, "A&C" = 1, "B&C" = 1,
                         "A&B&C" = 2),
                       shape = "ellipse")
plot(one_contained,
     quantities = list(type = "percent"))
```

### Russian doll

Sets intersecting inside other sets. This example also exemplifies how
the automatic label adjustment algorithm (borrowed from the excellent
**ggrepel** package) works.

```{r}
russian_doll <- euler(c("A" = 15, "B" = 0, C = 0, 
                        "A&B" = 10, "A&C" = 0, "B&C" = 0,
                        "A&B&C" = 5))
plot(russian_doll,
     labels = c(
       "A long and complicated\nlabel that is likely to have to be repelled",
       "Another, very\ntall label that\nprobably will intrude\ninto other\nlabels",
       "The spectacle before us\nwas indeed sublime")
     )
```

### Wilkinson set relationship

This set relationship is taken from Wilkinson et al. It works best
with ellipses.

```{r}
wilkinson <- euler(c(A = 4, B = 6, C = 3, D = 2, E = 7, F = 3,
                     "A&B" = 2, "A&F" = 2, "B&C" = 2, "B&D" = 1,
                     "B&F" = 2, "C&D" = 1, "D&E" = 1, "E&F" = 1,
                     "A&B&F" = 1, "B&C&D" = 1),
                   shape = "ellipse")
plot(wilkinson, 
     labels = list(fontfamily = "serif"),
     edges = list(lty = 3),
     quantities = list(type = "percent", 
                       font = 3))
```

### Gene set

```{r}
genes <- euler(c("SE" = 13, "Treat" = 28, "Anti-CCP" = 101, "DAS28" = 91,
                 "SE&Treat" = 1, "SE&DAS28" = 14, "Treat&Anti-CCP" = 6,
                 "SE&Anti-CCP&DAS28" = 1))
plot(genes,
     quantities = list(type = c("percent", "counts")))
```

### Three sets intersecting inside a fourth

```{r}
three_inside_fourth <- euler(c("A" = 30,
                               "A&B" = 3, "A&C" = 3, "A&D" = 3,
                               "A&B&C" = 2, "A&B&D" = 2, "A&C&D" = 2,
                               "A&B&C&D" = 1))
plot(three_inside_fourth)
```


### eulerAPE combination

A combination taken from the eulerAPE article.

```{r}
eulerape <- euler(c("a" = 3491, "b" = 3409, "c" = 3503,
                    "a&b" = 120, "a&c" = 114, "b&c" = 132,
                    "a&b&c" = 126),
                  shape = "ellipse",
                  control = list(extraopt = FALSE))
plot(eulerape)
```

### Four uniform interactions

```{r}
uniform <- euler(c("A" = 10, "B" = 10, "C" = 10, "D" = 10,
                   "A&B" = 3, "A&C" = 3, "A&D" = 0, "B&C" = 0, "B&D" = 3, "C&D" = 3,
                   "A&B&C" = 1, "A&B&D" = 1, "A&C&D" = 1, "B&C&D" = 1,
                   "A&B&C&D" = 1))
plot(uniform, labels = list(labels = c("Frodo", "Sam", "Merry", "Pippin"),
                            font = 1:4,
                            col = 1:4,
                            cex = seq(1, 1.5, length.out = 4)))
```


### Two circles intersecting completely

```{r}
two_overlapping <- euler(c("A" = 0, "B" = 0, "A&B" = 10))
plot(two_overlapping)
```

