---
title: "An introduction to eulerr"
author: "Johan Larsson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: eulerr.bib
vignette: >
  %\VignetteIndexEntry{An introduction to eulerr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**eulerr** generates area-proportional euler diagrams that display set
relationships (intersections, unions, and disjoints) with circles or ellipses.
[Euler diagrams](https://en.wikipedia.org/wiki/Euler_diagram) are Venn
diagrams without the requirement that all set interactions be present (whether
they are empty or not). That is, depending on input, eulerr will sometimes
produce Venn diagrams but sometimes not.

## Background

R features a number of packages that produce euler and/or venn diagrams;
some of the more prominent ones (on CRAN) are

* [eVenn](https://cran.r-project.org/package=eVenn),
* [VennDiagram](https://cran.r-project.org/package=VennDiagram),
* [venn](https://cran.r-project.org/package=venn),
* [colorfulVennPlot](https://cran.r-project.org/package=colorfulVennPlot), and
* [venneuler](https://cran.r-project.org/package=venneuler).

The last of these (venneuler) serves as the primary inspiration for this
package, along with the refinements that Ben Fredrickson has presented on his
[blog](http://www.benfrederickson.com/) and made available in his javascript
[venn.js](https://github.com/benfred/venn.js). The
[eulerAPE](http://www.eulerdiagrams.org/eulerAPE/) software, which was the
first program to uses ellipses instead of circles, has also been instrumental
in the design of **eulerr**. Whereas **eulerAPE** only handles 3 sets (that
all need to intersect), however, **eulerr** can (theoretically) take any number
of sets.

As for **venneuler** -- it has been known to produce 
[imperfect output](http://benfred.github.io/venn.js/tests/venneuler_comparison/)
for set configurations that have perfect solutions.

## Enter eulerr

**eulerr** is based on the improvements to **venneuler** that Ben Fredrickson
introduced with **venn.js** but has been programmed from scratch, uses different
optimizers, and returns statistics featured in **venneuler** and **eulerAPE** as
well as allows a range of different inputs and conditioning on additional 
variables.

### Input

At the time of writing, it is possible to provide input to `eulerr` as either

* a named numeric vector with set combinations as disjoint set combinations
or unions (depending on how the argument `type` is set),
* a matrix or data frame of logicals with columns representing sets and rows the
set relationships for each observation, or
* a list of sample spaces.

```{r input}
library(eulerr)
options(digits = 4)

# Input in the form of a named numeric vector
fit1 <- euler(c("A" = 25, "B" = 5, "C" = 5,
                "A&B" = 5, "A&C" = 5, "B&C" = 3,
                "A&B&C" = 3))

# Input as a matrix of logicals
set.seed(1)
mat <- cbind(
  A = sample(c(TRUE, TRUE, FALSE), size = 50, replace = TRUE),
  B = sample(c(TRUE, FALSE), size = 50, replace = TRUE),
  C = sample(c(TRUE, FALSE, FALSE, FALSE), size = 50, replace = TRUE)
)
fit2 <- euler(mat)
```

### Fit

We inspect our results by printing the eulerr object

```{r fit1_print}
fit2
```

or directly access and plot the residuals.

```{r residual_plot, fig.cap='Residuals for the eulerr fit.'}
# Cleveland dot plot of the residuals
dotchart(resid(fit2))
abline(v = 0, lty = 3)
```

This shows us that the `A&C` intersection is somewhat overrepresented in
`fit2`. Given that these residuals are on the scale of the original
values, however, the residuals are arguably of little concern.

As an alternative, we could plot the circles in another program by retrieving
their coordinates and radii.

```{r coefs}
coef(fit2)
```

### Starting configuration

A starting configuration is obtained via a constrained version of 
multidimensional scaling that has been explained thoroughly
[elsewhere](http://www.benfrederickson.com/better-venn-diagrams/). 

### Optimization

The starting configuration is based solely on the two-way relationships of the
sets so has to be optimized for most set relationships. We try to optimize the 
coordinates and radii of the solution with the objective of producing a diagram
that is as *accurate* as possible. In this context, however, accuracy is an
ambigious objective that has produced a slew of proposals. eulerr uses
the sums of squares as the target loss function to minimize 

$$\sum_{i=1}^{n} (y_i - \hat{y}_i) ^ 2$$
where $\hat{y}$ is the fitted disjoint areas.

### Goodness-of-fit

For goodness-of-fit measures, we the stress statistic from venneuler
[@wilkinson_2012]

$$\frac{\sum_{i=1}^{n} (y_i - \hat{y}_i) ^ 2}{\sum_{i=1}^{n} y_i ^ 2}$$
where $\hat{y}_i$ is an ordinary least squares estimate from the regression of
the fitted areas on the original areas that is being explored during
optimization.

We also provide the diagError statistic from eulerAPE (@micallef2014):

$$ \max_{i = 1, 2, \dots, n} \left| \frac{y_i}{\sum y_i} - \frac{\hat{y}_i}{\sum \hat{y}_i} \right|$$
In our example, the diagError is $`r fit2$diag_error`$ and our
stress is $`r fit2$stress`$, suggesting that the fit is accurate.

We can now be confident that eulerr provides a reasonable representation of
our input. Were it otherwise, we would do best to stop here and look for another
way to visualize our data. (I suggest the excellent
[UpSetR](https://cran.r-project.org/package=UpSetR) package.)

### Plotting

No we get to the fun part: plotting our diagram. This is easy, as well as
highly customizable, with **eulerr**

```{r eulerr_plot, fig.width=4, fig.height=4, fig.show='hold', fig.cap='eulerr plots can be modified in every possible way.', dev.args=list(type='cairo-png')}
plot(fit2)

# Remove fills, vary border type, and switch fontface.
plot(fit2, fill = "transparent", lty = 1:3, fontface = 4)
```

Plotting is provided through the excellent
[lattice](https://CRAN.R-project.org/package=lattice) and interfaces all of its
bells and whistles to allow for extensive customization of the resulting plots.

eulerr's default color palette is taken from
[qualpalr](https://cran.r-project.org/package=qualpalr) -- another
package that I have developed -- which uses color difference algorithms to 
generate distinct qualitative color palettes.

## Acknowledgements

eulerr would not be possible without Ben Fredrickson's work on
[venn.js](http://www.benfrederickson.com) or Leland Wilkinson's
[venneuler](https://cran.r-project.org/package=venneuler).

## References
