---
title: "loss-functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{loss-functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
```

eulerr features multiples loss functions, which result in different diagrams
for many combinations. In this vignette, we visualize the effect of the loss
function on an example from an issue [posted on the GitHub repository for
eulerr](https://github.com/jolars/eulerr/issues/93).

We list the combinations below, which consists of 5 different sets,
*agc*, *camk*, *cmgc*, and *tk*.

```{r}
combos <- c(
  "agc"                  = 9,
  "camk"                 = 17,
  "cmgc"                 = 16,
  "tk"                   = 16,
  "tkl"                  = 23,
  "agc&camk"             = 1,
  "camk&tk"              = 1,
  "tk&tkl"               = 1,
  "camk&cmgc&tkl"        = 1,
  "camk&tk&tkl"          = 2,
  "agc&camk&tk&tkl"      = 1,
  "camk&cmgc&tk&tkl"     = 3,
  "agc&camk&cmgc&tk&tkl" = 1
)
```

Notice that the sizes of most of the intersections are small compared to the
size of the sets themselves and that many of the
intersections are missing. Generating an exact Euler diagram that shows these
intersections and at the same time omits the intersections that are here
implicitly 0 is an impossible problem, which means that the best
we can do is an approximation.

What kind of approximation we get depends on the loss functions we use.
If we use the default, which in **eulerr** is the sums of squared errors, we
will almost certainly get a design in which the intersections involving many
sets are missing since including them inevitably leads to larger errors from
having to include other intersections that are not present.

```{r fig.cap = "An Euler diagram fit to the combination given earlier on, showing that only 1-by-1 intersections are present. This fit uses the default loss function, the sum of squared errors."}
library(eulerr)

fit <- euler(combos)
plot(fit)
```

If we rather want a diagram that includes these intersections, despite 
leading to errors for the zero-intersections, then we need to switch the loss
function we use. In **eulerr** there are currently five different
loss metrics, which can be provided to the `loss` argument in `euler()`. 

Table: Loss functions in **eulerr**

Loss                              Input          Definition
----------------------            ------------   -----------------------------
Sum of squared errors             `sum_sq`       $\sum_i (y_i - \hat y_i)^2$
Sum of absolute errors            `sum_abs`      $\sum_i |y_i - \hat y_i|$
Sum of regionErrors               `sum_reg`      $\sum_i \big|y_i/\sum_k y_k - \hat y_i / \sum_k \hat y_k \big|$
Maximum squared error             `max_sq`       $\max_i (y_i - \hat y_i)^2$
Maximum absolute error            `max_sq`       $\max_i |y_i - \hat y_i|$
Maximum regionError (diagError)   `sum_reg`      $\sum_i \big|y_i/\sum_k y_k - \hat y_i / \sum_k \hat y_k \big|$

To see what these different choices mean for the combination that we have
looked at, we now refit the diagram for each combination.

```{r fig.cap = "Euler diagrams fit to the combination above, using different loss function", fig.show = "hold"}
metrics <- c(
  "sum_sq",
  "sum_abs",
  "sum_reg",
  "max_sq",
  "max_abs",
  "max_reg"
)

for (loss in metrics) {
  fit <- euler(combos, loss = loss)
  print(plot(fit, main = loss))
}
```

As you can see, the errors that sum either the absolute or squared errors
result in very similar fits and keep the existing two-set intersections and
drop everything else. The `max_abs` and `max_sq` errors, meanwhile, produce
fits that are much more unpredictable since they only care about the largest
error. Finally, `max/sum_reg` results diagrams that tries to include many more
intersections.

Feel free to raise a request at <https://github.com/jolars/eulerr/issues>
if you know of any other loss function that you think should be included
in the package.
