---
title: "eulerr under the hood"
author: "Johan Larsson"
date: "`r Sys.Date()`"
output: pdf_document
toc: true
indent: true
colorlinks: true
papersize: a4
fontfamily: libertine
geometry:
  - margin=3.5cm
bibliography: eulerr.bib
vignette: >
  %\VignetteIndexEntry{eulerr under the hood}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::knit_hooks$set(crop = knitr::hook_pdfcrop)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 4,
  fig.height = 4,
  crop = TRUE
)

options(digits = 3)

library(lattice)
lattice.options(default.theme = list(fontsize = list(text = 10, points = 5)))

set.seed(1)
```

# Introduction

**eulerr** relies on an extensive machinery to turn user input into a pretty
Euler diagram. By design, little, or in fact none, of this ever requires any
tinkering from the user. The downside of this, however, is that it
requires eulerr to make several well-formed decisions about the 
design of the diagram on behalf of the user.

# Input

To start with, we need input in the form of

* a named numeric vector, such as `c(A = 10, B = 5, "A&B" = 3)`, where 
ampersands define disjoint set combinations or unions, depending on the
argument `input`,
* a data.frame or matrix of logicals or binary indices where each row denotes
the set relationships of
  - either a single observation
```{r}
matrix(sample(c(TRUE, FALSE), 12, replace = TRUE),
       ncol = 3,
       dimnames = list(NULL, c("A", "B", "C")))
```
  - or of a unique set combination if a numeric vector is supplied to the
    argument `weights`,
```{r}
matrix(c(TRUE, FALSE, FALSE,
         TRUE, TRUE, FALSE,
         FALSE, FALSE, TRUE),
       ncol = 3,
       dimnames = list(NULL, c("A", "B", "C")))
```
* a table (max 3 dimensions), 
```{r}
as.table(apply(Titanic, 2:4, sum))
```
* or a list of sample spaces, such as
```{r}
list(A = c("x", "xy", "xyz"),
     B = c("xy"),
     C = c("x", "xyz"))
```

If the data.frame or matrix form is used, the user additionally has the option
to split the data set by a factor and compute separate euler diagrams for 
each split. This is accomplished by supplying a factor variable to the 
`by` arguments (see the documentation in `?base::by`).

# Pre-processing

eulerr begins by organizing the input of the user into a matrix of binary
indexes, which in R is represented as a matrix of logicals. For a three set 
configuration, this looks like this,

```{r}
library(eulerr)
eulerr:::bit_indexr(3)
```

and is accompanied by a vector of the *disjoint* areas of the set combinations.

Given these areas, we figure out the required pairwise distance between
the sets (as circles) to achieve a circle--circle overlap that matches the
set intersection between the sets. We do this numerically, using the 
formula for the circle--circle overlap,

\begin{multline}
A = r_1^2\arccos\left(\frac{d^2 + r_1^2 - r_2^2}{2dr_1}\right) + 
r_2^2\arccos\left(\frac{d^2 + r_2^2 - r_1^2}{2dr_2}\right) - \\
\frac{1}{2}\sqrt{(-d + r_1 + r_2)(d + r_1 - r_2)(d - r_1 + r_2)(d + r_1 + r_2)},
\end{multline}

where *r~1~* and *r~2~* are the radii of the first and second circles
respectively and *d* the distance between the circles.

*r~1~* and *r~2~* are known but because *d* is not, we attempt to find it
using one-dimensional numerical optimization. As our loss function we use
the squared difference between *A* and the desired overlap and then optimize
using **R**'s `optimize()`. This could look something like this.

```{r}
r1 <- 0.7 #radius of set 1
r2 <- 0.9 #radius of set 2
overlap <- 1 #area of overlap

stats::optimize(eulerr:::discdisc, #computes the squared loss mentioned above
                interval = c(abs(r1 - r2), sum(r1, r2)),
                r1 = r1,
                r2 = r2,
                overlap = overlap)

# minimum is our required distance
```

Now that we have the distances, we can proceed to the next step: computing
an initial layout.

# Initial configuration

The next step is to provide a starting layout. This can be done in a number of 
ways; eulerr uses the primary method from Ben Fredrickson's
[venn.js](https://github.com/benfred/venn.js/) script -- basically a constrained
version of multi-dimensional scaling (MSD), which is also used in Leland
Wilkinson's [venneuler](https://CRAN.R-project.org/package=venneuler) package
[@wilkinson_2012]. However, whereas venneuler's version tries to place
disjoint sets exactly neck-in-neck and completely contained sets at the precise
midpoints of their parents, the version from venn.js (and eulerr) accepts any
disjoint or contained positions as long as they are outside or inside of the
parent sets respectively.

The MDS algorithm works by trying to optimize the pairwise relationsships
between sets using the following loss and gradient functions

> \[\small
> loss = \sum_i \sum_j { {\begin{cases}
>     0 & \text{disjoint}(i, j)\\ 
>     0 & \text{subset}(i, j)\\ 
>     ((X_{i} - X_{j})^T(X_{i} - X_{j}) - D_{ij}^2) ^2  & \text{otherwise} \\ 
> \end{cases}}}
> \]
>
> \[\small
> \nabla f(X_{i}) = \sum_j {\begin{cases}
>      \vec{0} & \text{disjoint}(i, j)\\ 
>      \vec{0} & \text{subset}(i, j)\\ 
>      4 {((X_{i} - X_{j})^T(X_{i} - X_{j}) - D_{ij}^2)} (X_{i} -
>      X_{j}) & \text{otherwise} \\ 
> \end{cases}}\]
> *source: [Better Venn Diagrams](http://www.benfrederickson.com/better-venn-diagrams/) by Ben Fredrickson,
> which also includes a nice interactive demonstration of how it works*.

Fredrickson uses the Polak--Ribière Conjugate Gradient Method to optimize
the initial layout. In my experience this method occasionally ends
up in local minima, which is why
I have opted to use `nlm()` from the **R** core package `stats`, which is a 
translation from FORTRAN code developed by Schnabel et al.
[@schnabel_1985] and uses a mixture of algorithms (Newton and 
Quasi-Newton) to derive at a solution.

This initial configuration will work perfectly for any 1--2 set combinations
and as well as possible with 3 sets if we use circles but for all other
combinations there is usually a need to fine tune the configuration.

# Final configuration

In order to finalize the configuration we need to be able to compute the
areas of the overlaps of the sets, which as it turns out, is *not* trivial. In
fact, most of methods rely on approximations of the areas by, for instance,
quad-tree binning (venneuler) or polygon intersections
(VennMaster [@kestler_2008]). These methods yield reasonable
estimates but, given that the computation may have to run for a vast number
of iterations, are usually prohibitive in terms of performance.

venn.js and eulerAPE both, however, use exact algorithms. Based on the fact
that any intersection of circles or ellipses can be represented
as a convex polygon with elliptical segments on the fringes, it is possible
to arrive at exact area calculations. 

## Intersections

Finding the areas of the overlaps exactly requires that we first know the points
at which the different ellipses intersect. eulerr's approach to this is based
on a method outlined by Richter--Gebert in
*Perspectives on Projective Geometry* [@richter-gebert_2011].
eulerr owes significant debt to the **R** package *RConics*
[@huber_2014], which has been tremendously helpful
in developing and, especially, debugging the algorithm. Some parts of the code
are in fact straight-up translations to C++ from the code in *RConics*.

The method is based in *projective geometry* (rather than euclidean). To find
the intersection points, the algorithm first

* converts the two ellipses from canonical form to matrix notation. The 
canonical form of a rotated ellipse is given by
\[
\frac{((x-h)\cos(\phi)+(y-k)\sin(\phi))^2}{a^2}+\frac{((x-h) \sin(A)-(y-k) \cos(\phi))^2}{b^2} = 1,
\]
where *phi* is the counter-clockwise angle from the positive x axis to the 
semi-major axis *a*. *b* is the semi-minor axis whilst *(h, k)* is the center
of the ellipse. This is then converted to the matrix form
\[
E = \begin{bmatrix}A & B/2 & D/2 \\
                   B/2 & C & E/2 \\
                   D/2 & E/2 & F
    \end{bmatrix},
\]
which may be used to represent any conic. We then
* split one of the ellipses (conics) into a pencil of two lines, and subsequently
* intersect the remaining conic with these two lines, which will yield between
0 and 4 intersection points.

[TODO]: # Add some illustrations of this.

## Areas

The next step is to calculate the area of overlap between all the
possible combinations of ellipses. The solution to this was discovered,
as far as I know, by Ben Fredrickson who explains it beautifully in a 
[blog post](http://www.benfrederickson.com/calculating-the-intersection-of-3-or-more-circles/).
It relies on finding all the intersection points between the currently
examined sets that are also within these sets. It is then trivial to find the
area of the convex polygon that these vertices make up. Finding the 
rest of the area, which is made up of the ellipse segments between
subsequent points, requires a bit of trigonometry. 

Here, we have used an algorithm from David Eberly [@eberly_area_2016], which
computes circle integral between the points on the ellipse minus the 
area of the triangle made up of the center of the ellipse: 
\[
A(\theta_0, \theta_1) = F(\theta_1) - F(\theta_1) - \frac{1}{2}|x_1y_0 - x_0y_1|,
\]
\[
\text{where } F(\theta) = \frac{a}{b}\left[ \theta -
\arctan{\left(\frac{(b - a)\sin{2\theta}}{b + a +(b - a )\cos{2\theta}} \right)} \right]
\]

As our loss function, we use the sum of squared differences between the 
disjoint set intersections and the areas we have computed and again use the `nlm()`
optimizer to layout the set. 

This optimization step is the bottleneck of the overall computations in terms of
performance, being that 
we're optimizing over 5 parameters for every ellipse (or 3 in the case
of circles) -- nevertheless, we're profitting immensely from the implementation
in the C++ programming language through **Rcpp** [@eddelbuettel_2011] and its
plugin for the linear algebra library **Armadillo** [@eddelbuettel_2014] which
ends up making the code much faster than the java-based venneuler.

# Layout

Since the optimization steps are unconstrained, we run the risk of
ending up with dispersed layouts. To fix this, we use the SKYLINE-BL 
rectangle packing algorithm [@jylanki_2010] to pack the disjoint clusters of
ellipses (in case there are any) into a heuristically chosen bin.

At the time of writing this algorithm is crudely implemented -- for 
instance, it does not attempt to rotate the rectangles (boundaries for the
ellipses) or attempt to use. Since we're dealing with a rather simple version
of the rectangle packing problem, however, it seems to do the trick.

# Output

Before we get to plotting the solution, it is useful to know how well the
fit from eulerr matches the input. Sometimes euler diagrams are just not
feasible, particular for combinations with many sets, in which case we should
stop here and look for another design to visualize the set relationships.

It is not, however, obvious what it means for a euler diagram to "fit well". 
venneuler uses a metric called *stress*, which is defined as
\[
\frac{\sum_{i=1}^{n} (y_i - \hat{y}_i) ^ 2}{\sum_{i=1}^{n} y_i ^ 2}
\]
where $\hat{y}_i$ is an ordinary least squares estimate from the regression of
the fitted areas on the original areas that is being explored during
optimization.

Meanwhile, eulerAPE [@micallef2014] uses *diagError*:
\[
\max_{i = 1, 2, \dots, n} \left| \frac{y_i}{\sum y_i} - \frac{\hat{y}_i}{\sum \hat{y}_i} \right|
\]

Both metrics are given the user after the diagram has been fit, together
with a table of residuals.

```{r, fig.cap = "A plot with circles."}
combo <- c("A" = 1, "B" = 1, "C" = 1,
           "A&B" = 0.5, "A&C" = 0.5, "C&B" = 0.5)

fit1 <- euler(combo)
fit1
plot(fit1, counts = TRUE)
```

It is clear that this is not a good fit, which we can find out just by
looking at the plot. This is a good example of when ellipses come in handy.

```{r, fig.cap = "A plot with ellipses."}
fit2 <- euler(combo, shape = "ellipse")
fit2
plot(fit2, counts = TRUE)
```

Much better.

[TODO]: # What is a high and low diag_error?

# Plotting

Let's face it: euler diagrams are naught without visualization. Here, eulerr
interfaces the elegant lattice graphics system [@sarkar_2008] to grant the
user extensive control over the output, and allow for facetted plots in
case such a design was used in fitting the euler configuration.

## Labelling
Most users will want to label their euler diagrams. One option is to simply 
add a legend

```{r, fig.cap = "A simple plot."}
plot(euler(c(A = 2, B = 3, "A&B" = 1)), auto.key = TRUE)
```

but many will want to label their diagrams directly, perhaps with counts.

```{r, fig.cap = "A plot with counts."}
plot(euler(c(A = 2, B = 3, "A&B" = 1)), counts = TRUE)
```

In this case, layout out the diagram becomes considerably more involved. Finding
a reasonable spot for the text inside the diagram only lends itself to an
easy solution if the shape of the intersection has a center-of-gravity inside
ellipse, in which case an average of some of the points might suffice. This
is often not the case, however, and we need a better solution. Specifically,
what we need is a method to find the point inside the circle overlap for the
counts and circle complement to the intersection for our labels.

So far, we have not been able to derive at an analyitcal solution for
finding a good point, or for that matter a reliable way of finding *any* point
that is in the required intersection or complement. As is often the case,
the next-best thing turns out to be a numerical one. First, we locate a 
point that is inside the required region by spreading points across one of the
discs involed in the set combination. To spread points uniformly, we use
*Vogel's method* [@arthur_2015, @vogel_1979]
\[
\left( p_k = (\rho_k, \theta_k) = \left( r \sqrt{\frac{k}{n}},\, \pi (3 - \sqrt{5})(k - 1) \right) \right)_{k=1}^n,
\]
which is actually based on the golden angle.

```{r, fig.cap = "Spreading points on a disc with Vogel's method."}
n <- 500
seqn <- seq(0, n, 1)
theta <- seqn*pi*(3 - sqrt(5))
rad <- sqrt(seqn/n)
x <- rad*cos(theta)
y <- rad*sin(theta)

plot(x, y, asp = 1, pch = 16, axes = FALSE, xlab = "", ylab = "")
```

After this, we scale, translate,
and rotate the points so that they fit the desired ellipse.

[TODO]: # Visualize this step.

After we've spread our points throughout the ellipse and found one that
matches our desired combination of ellipses/sets, we then proceed to optimize
its position numerically. For this, we use version of the
Nelder--Mead method [@nelder_1965] which we've translated from Matlab
code by Kelley [@kelley_1999] and customized for eulerr
(in particular to make sure that
the simplex does not escape the intersection boundaries since we for this
problem *want* the local minimum).

## Coloring

Per default, the ellipses are filled with colors. The default option is to
use an adaptive scheme in which colors are chosen to provide a balance
between dinstinctiveness, beauty, and consideration for the color deficient.
The color palette has been generated from
[qualpalr](https://CRAN.R-project.org/package=qualpalr)
(developed by the author), which automatically generates qualitative color
palettes based on a model of color perception.

# References
